### 2021-05-12 Supplementary_Material/Missing_data.Rmd
Pollen nectar pPCA added

########################################################################################################################################################
#SCRIPT TO CHECK PATTERNS OF MISSING VALUES
########################################################################################################################################################
#1) READ TRAIT DATA

#2) CLEAN DATA

#3) EXPLORE PATTERNS OF MISSING DATA
########################################################################################################################################################
#LOAD LIBRARIES
library(readxl) #read excel file (trait data)
library(dplyr) #data manipulation
library(missMDA) #For missing values
library(FactoMineR) #Produce pca biplot with individuals and variables
library(factoextra)
library(ggpubr) 
library(missForest) #random forst imputation
library(rtrees) #for phylogenetic distancelibrary(MASS)
library(ape)
library(PVR) #calculate eigen vectors and add them as a column in the imputation, seems to improve output
library(visdat)
library(tidyverse)

########################################################################################################################################################
#1) READ TRAIT DATA
########################################################################################################################################################
#load data
trait_data <- read_excel("Data/Trait_data_raw/Trait_data_final.xlsx",na = "NA")
########################################################################################################################################################
#2) A) CLEAN DATA AND B) RENAME LEVELS FOR IMPUTATION 
########################################################################################################################################################
#####
#A) CLEAN DATA
#####
#select just filled rows
trait_data_1 <- trait_data[1:1712,] #It may be a more elegant way but this does the job, 1712 rows of data

#str(trait_data_1)
#filter data, select species with flower level info and capitulum
trait_filtered <- filter(trait_data_1, Info_level == "flower" |  Info_level == "capitulum")
#levels(as.factor(trait_filtered$Info_level)) #checking levels
#remove NA's and duplicated species|Some spp are repeated once the spp names are standardize with taxize
trait_filtered$Species_all[trait_filtered$Species_all=="NA"]<-NA 
trait_filtered_1 <- trait_filtered[!is.na(trait_filtered$Species_all),]
trait_filtered_2 <- trait_filtered_1[!duplicated(trait_filtered_1$Species_all),]
#str(trait_filtered_2)

#Select columns to work with
t <- trait_filtered_2[c("Species_geonet","Family_all","Genus_all","Species_all","Breeding_system","Compatibility_system",
                        "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
                        "Flowers_per_plant", "Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
                        "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul")]

####################################################################################################
#2) CLEAN DATA. I REPEAT THE SAME PROCEDURE OF THE DATA IMPUTATION TO GET THE SAME NUMBER OF SPECIES
####################################################################################################
#Before data imputation I need to calculate the phylo of the species
#remove not found species that do not run with get tree
cols.num <- c("Family_all","Genus_all","Species_all")
t[cols.num] <- sapply(t[cols.num],as.character)
t$Species_all <- gsub("Species_all_", "", t$Species_all)

@@

#remove species that are not until species level
#ALL THE SPECIES WITH SP. ARE DELETD
t$Species_geonet <- gsub("M_PL_.*","",t$Species_geonet) #subsitute partial string for nothing
t$Species_geonet <- gsub(" $","", t$Species_geonet, perl=T) #remove trailing spaces
t$Species_geonet <-  sub('sp.1$', 'sp.', t$Species_geonet)#PREPARE ALL SP TO SP.
t$Species_geonet <- sub('sp.2$', 'sp.', t$Species_geonet)#PREPARE ALL SP TO SP.
t$Species_geonet <- sub('sp$', 'sp.', t$Species_geonet) #PREPARE ALL SP TO SP.
t$Species_geonet <- sub("sp.$", "DELETE", t$Species_geonet) #change all sp. for DELETE
t <- t[- grep("DELETE",  t$Species_geonet),] #remove species with "DELETE"
#CHECK LEVELS
#levels(factor(t$Species_all))

@@

#nrow(t)

#Select just columns with traits
t <- t[c("Breeding_system","Compatibility_system",
                        "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
                        "Flowers_per_plant", "Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
                        "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul")]

@@

colnames(t) <- c("Breeding system","Compatibility system",
         "Autonomous selfing qualitative","Autonomous selfing quantitative", "Flower shape", "Flower symmetry", 
         "Flower number per plant", "Inflorescence width", "Flower width", "Flower length","Style length", "Ovule number", 
         "Life form", "Lifespan","Plant height","Nectar qualitative","Nectar quantitative")

>>>>

########################################################################################################################################################
#SCRIPT FOR DATA IMPUTATION
########################################################################################################################################################
#1) READ TRAIT DATA

#2) CLEAN DATA AND RENAME LEVELS

#3) EXPLORE PATTERNS OF MISSING DATA

#4) CALCULATE PHYLOGENETIC DISTANCE TO CORRECT WITH EIGENVALUES IN THE IMPUTATION

#5) ICLUDE EIGENVALUES IN RAWDATA TO IMPROVE IMPUTATION OURPUT

#6) IMPUTE DATA

#7) SAVE IMPUTED DATA 
########################################################################################################################################################
#LOAD LIBRARIES
library(readxl) #read excel file (trait data)
library(dplyr) #data manipulation
library(missMDA) #For missing values
library(FactoMineR) #Produce pca biplot with individuals and variables
library(factoextra)
library(ggpubr) 
library(missForest) #random forst imputation
library(rtrees) #for phylogenetic distancelibrary(MASS)
library(ape)
library(PVR) #calculate eigen vectors and add them as a column in the imputation, seems to improve output
library(visdat) #VISUALIZE MISSING DATA
library(naniar)
library(tidyverse)
########################################################################################################################################################
#1) READ TRAIT DATA
########################################################################################################################################################
#load data
trait_data <- read_excel("Data/Trait_data_raw/Trait_data_final.xlsx",na = "NA")
########################################################################################################################################################
#2) A) CLEAN DATA AND B) RENAME LEVELS FOR IMPUTATION 
########################################################################################################################################################
#####
#A) CLEAN DATA
#####
#select just filled rows
trait_data_1 <- trait_data[1:1712,] #It may be a more elegant way but this does the job, 1712 rows of data

str(trait_data_1)
#filter data, select species with flower level info and capitulum
trait_filtered <- filter(trait_data_1, Info_level == "flower" |  Info_level == "capitulum")
levels(as.factor(trait_filtered$Info_level)) #checking levels
#remove NA's and duplicated species|Some spp are repeated once the spp names are standardize with taxize
trait_filtered$Species_all[trait_filtered$Species_all=="NA"]<-NA 
trait_filtered_1 <- trait_filtered[!is.na(trait_filtered$Species_all),]
trait_filtered_2 <- trait_filtered_1[!duplicated(trait_filtered_1$Species_all),]
str(trait_filtered_2)

#Select columns to work with
t <- trait_filtered_2[c("Species_geonet","Order_all","Family_all","Genus_all","Species_all","Breeding_system","Compatibility_system",
                        "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
                        "Flowers_per_plant", "Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
                        "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul","Nectar_mg","Nectar_concentration",
                        "Pollen_per_flower")]

########################################################################################################################################################
#####
#B) RENAME LEVELS FOR IMPUTATION
#####
#Breeding_system| For the purpose of this study we are not going to consider other complex breeding systems and just focus on 
#Hermaphroditism, dioecy and monoecy, the closest levels are grouped within each category
t$Breeding_system <- as.character(t$Breeding_system)
t$Breeding_system[t$Breeding_system=="androdioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="androgynodioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="andromonoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="gynodioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="gynoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="gynomonoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="gynomonoecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="monoecious_dioecious"] <- "monoecious"
t$Breeding_system[t$Breeding_system=="polygamo-dioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="protandrous"] <- "hermaphrodite"
t$Breeding_system[t$Breeding_system=="protogynous"] <- "hermaphrodite"
t$Breeding_system[t$Breeding_system=="subdioecious"] <- "dioecious"
t$Breeding_system[t$Breeding_system=="submonoecious"] <- "monoecious"
#Change to capital letters| There are already levels with lower case
t$Breeding_system[t$Breeding_system=="monoecious"] <- "Monoecious"
t$Breeding_system[t$Breeding_system=="dioecious"] <- "Dioecious"
t$Breeding_system[t$Breeding_system=="hermaphrodite"] <- "Hermaphrodite"
levels(as.factor(t$Breeding_system))
t$Breeding_system <- as.factor(t$Breeding_system)

#Now 3 levels
#Flower_morphology
t$Flower_morphology <- as.character(t$Flower_morphology)
t$Flower_morphology[t$Flower_morphology=="bowl"] <- "open"
t$Flower_morphology[t$Flower_morphology=="dish"] <- "open"
t$Flower_morphology[t$Flower_morphology=="exposed"] <- "open"
t$Flower_morphology[t$Flower_morphology=="spadix"] <- "spike"
t$Flower_morphology[t$Flower_morphology=="open"] <- "Open"
t$Flower_morphology[t$Flower_morphology=="brush"] <- "Brush"
t$Flower_morphology[t$Flower_morphology=="campanulate"] <- "Campanulate"
t$Flower_morphology[t$Flower_morphology=="capitulum"] <- "Capitulum"
t$Flower_morphology[t$Flower_morphology=="funnelform"] <- "Funnelform"
t$Flower_morphology[t$Flower_morphology=="papilionaceous"] <- "Papilionaceous"
t$Flower_morphology[t$Flower_morphology=="spike"] <- "Spike"
t$Flower_morphology[t$Flower_morphology=="tube"] <- "Tube"
levels(as.factor(t$Flower_morphology))
t$Flower_morphology <- as.factor(t$Flower_morphology)

#Life span
t$lifespan <- as.character(t$lifespan )
t$lifespan[t$lifespan=="annual"] <- "short_lived"
t$lifespan[t$lifespan=="biennial"] <- "short_lived"
t$lifespan[t$lifespan=="short_lived"] <- "Short lived"
t$lifespan[t$lifespan=="perennial"] <- "Perennial"
levels(as.factor(t$lifespan))
t$lifespan <- as.factor(t$lifespan )

########################################################################################################################################################
#3) EXPLORE PATTERNS OF MISSING DATA
########################################################################################################################################################
#Explore patterns of missing data
missing_data <- unlist(lapply(t, function(x) sum(is.na(x))))/nrow(t)
sort(missing_data[missing_data >= 0], decreasing=T)
#nectar values all above 50% except presence/absence
#then, just one cololumn with 68% missing data, then one with 35% and then 20% and 10%. The rest under 5%.
#it is recommended to remove columns over 50% but we are particulary interested in having numeric levels of quantitative selfing 
#we are going to work just with presence/absence of nectar
#we will perform further analysis just with a subset with the quantitative values o nectar (very interesting data and diffciult to obtain)

#We convert cualitative data to quantitative to reduce missing data on this column and hence conduct the imputation 
#use ifelse base r does not work with these NA'S Even with work around. base r does not like the na option of read excel
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("high") & is.na(t$Autonomous_selfing_level_fruit_set), 88, t$Autonomous_selfing_level_fruit_set)
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("medium") & is.na(t$Autonomous_selfing_level_fruit_set), 50.5, t$Autonomous_selfing_level_fruit_set)
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("low") & is.na(t$Autonomous_selfing_level_fruit_set), 13, t$Autonomous_selfing_level_fruit_set)
t$Autonomous_selfing_level_fruit_set <- ifelse(t$Autonomous_selfing_level %in% c("none") & is.na(t$Autonomous_selfing_level_fruit_set), 0, t$Autonomous_selfing_level_fruit_set)

#t$Nectar_ul adding it here just to see how many cells 
#Now we just select presence/absence of nectar
#select columns of interest
t <- t[c("Species_geonet","Order_all","Family_all","Genus_all","Species_all","Breeding_system","Compatibility_system",
         "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
         "Flowers_per_plant", "Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
         "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul","Nectar_mg","Nectar_concentration",
         "Pollen_per_flower")]

#check missing data now
missing_data <- unlist(lapply(t, function(x) sum(is.na(x))))/nrow(t)*100
sort(missing_data[missing_data >= 0], decreasing=T)

missing_data <- unlist(lapply(t, function(x) nrow(t)-sum(is.na(x))))
sum(missing_data)

#Number of missing values on the datset
#vis_miss(t[,c(6:20)])
#9.9% percentage of missing values]

#Checking number of filled cells/ just selecting variables used in the manuscript; nectar ul should be accounted too and it has been considered but
#not added to run this code

########################################################################################################################################################
#4) CALCULATE PHYLOGENETIC DISTANCE TO CORRECT WITH EIGENVALUES IN THE IMPUTATION
########################################################################################################################################################
#Convert to factor 
cols <- c("Order_all", "Family_all", "Genus_all", "Species_all", "Compatibility_system", "Autonomous_selfing_level", "Flower_morphology",
          "Flower_symmetry", "life_form", "lifespan","Nectar_presence_absence")
t[cols] <- lapply(t[cols], factor)  ## as.factor() could also be used
str(t)

#Impute data considering phylogenetic distance of species
#it improves quality of imputation
#https://doi.org/10.1111/2041-210X.12232 see ref for method comparisons with and without imputation

#Multiple imputation process with phylogeny --> Missforest
#https://doi.org/10.1111/2041-210X.12232
#althought it does not have an option we are going to add it as a column

#Before data imputation I need to calculate the phylo of the species
#remove not found species that do not run with get tree
cols.num <- c("Family_all","Genus_all","Species_all")
t[cols.num] <- sapply(t[cols.num],as.character)
t$Species_all <- gsub("Species_all_", "", t$Species_all)

@@

#remove species that are not until species level
#ALL THE SPECIES WITH SP. ARE DELETD
t$Species_geonet <- gsub("M_PL_.*","",t$Species_geonet) #subsitute partial string for nothing
t$Species_geonet <- gsub(" $","", t$Species_geonet, perl=T) #remove trailing spaces
t$Species_geonet <-  sub('sp.1$', 'sp.', t$Species_geonet)#PREPARE ALL SP TO SP.
t$Species_geonet <- sub('sp.2$', 'sp.', t$Species_geonet)#PREPARE ALL SP TO SP.
t$Species_geonet <- sub('sp$', 'sp.', t$Species_geonet) #PREPARE ALL SP TO SP.
t$Species_geonet <- sub("sp.$", "DELETE", t$Species_geonet) #change all sp. for DELETE
t <- t[- grep("DELETE",  t$Species_geonet),] #remove species with "DELETE"
#CHECK LEVELS
levels(factor(t$Species_all))

@@

#Select just columns with traits
t <- t[c("Breeding_system","Compatibility_system",
                        "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
                        "Flowers_per_plant", "Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
                        "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul","Nectar_mg","Nectar_concentration",
         "Pollen_per_flower")]

@@

colnames(t) <- c("Breeding system","Compatibility system",
         "Autonomous selfing qualitative","Autonomous selfing quantitative", "Flower shape", "Flower symmetry", 
         "Flower number per plant", "Inflorescence width", "Flower width", "Flower length","Style length", "Ovule number", 
         "Life form", "Lifespan","Plant height","Nectar qualitative","Nectar per flower (microlitres)", "Nectar per flower (miligrames)", "Nectar concentration (%)", "Pollen grains per flower")
